Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
Uniqueness of hylos

The remainder of the paper is structured as follows:
\begin{itemize}
  \item In Section~\ref{sec:rec-schemes} we give an overview about recursion
schemes.
  \item In Section~\ref{sec:containers} we formalise the type of container
functors ensuring the presence of least and greatest fixed-points for functors
and suitably adapted for program extraction.
  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
hylomorhisms and their universal properties.
  \item In Section~\ref{sec:examples} we use the framework to formalise examples
of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
Furthermore we verify the short-cut fusion optimisation and show the extracted
optimised code to OCaml.
\end{itemize}

@@

\subsubsection{Haskell Implementation of Hylomorphisms}
% which is very useful in list processing. As a fold, The frequently used function
% map is also a fold:
% \begin{minted}{haskell}
% map :: (a -> b) -> [a] -> [b]
% map f = foldr (\x xs -> (f x):xs) []
% \end{minted}
We implement the hylomorphism tightly following the theory explained above:
\begin{minted}{haskell}
hylo :: Functor f => (c -> f c) -> (f a -> a) -> c -> a
hylo c a = a . fmap (hylo c a) . c
\end{minted}

A classic example of a divide-and-conquer algorithm that can be written as an hylomorphism is the quicksort program below:

\begin{minted}{haskell}
partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r >= x])

quicksort :: [Int] -> [Int]
quicksort [] = []
quicksort (x:xs) = let (xs, x, ys) = partition x xs
                   in (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
The partition program takes an element \mintinline{haskell}{x} and a list and returns a triple consisting of a list with all the elements in the list smaller than \mintinline{haskell}{x}, the element itself and a list with all the elements in the list greater or equal than  \mintinline{haskell}{x}. The quicksort program first runs the partition program, which divides the list in three parts, calls itself recursively on the sublists and then puts the results together using the concatenation function.

To implement this as an hylo we define the functor representing trees where leaves are just empty and each node contains a value:
\begin{minted}{haskell}
data TreeF a k = Leaf | Branch k a k deriving Functor
\end{minted}
Now the partition program can be written as a  \mintinline{haskell}{TreeF}-coalgebra
\begin{minted}{haskell}
divide :: [Int] -> TreeF Int [Int]
divide [] = Leaf
divide (p:xs) = Branch [l | l <- xs, l < p] p [r | r <- xs, r > p]
\end{minted}
and the concatenation function can be written as  \mintinline{haskell}{TreeF}-algebra
\begin{minted}{haskell}
conquer :: TreeF Int [Int] -> [Int]
conquer Leaf = []
conquer (Branch ls x rs) = ls ++ [x] ++ rs
\end{minted}
Finally, quicksort can be rewritten as an hylomorphism
\begin{minted}{haskell}
quicksort = hylo divide conquer
\end{minted}
As an important remark we have used the in-built list type for keeping the code
as tidy as possible, but it would have been possible to use the inductive
\mintinline{haskell}{List} type we defined above and define both
\mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via
catamorphisms.

Finally, it is important to note that both  \mintinline{haskell}{divide} and
\mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on
\mintinline{haskell}{ListF}. These are rather (co)algebras on
\mintinline{haskell}{ListF}.  Hence the terminating argument relies solely on
the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there
is an unique solution to \mintinline{haskell}{hylo divide conquer}.

\subsubsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

\noindent
As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
hylo : forall F `{setoid A} `{setoid B}, Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from this uniqueness property:
\begin{coqcode}
Lemma hylo_fusion_l
  (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
  : f2 \o g1 =e g2 \o fmap f2 ->
    f2 \o hylo g1 h1 =e hylo g2 h1.

@@

apply hylo_univ.
       (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
rewrite fmap_comp.
       (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
rewrite [...]. (* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
rewrite <- E2.
       (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1      *)
rewrite [...]. (* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
rewrite <- hylo_unroll.
       (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1                             *)
\end{coqcode}
  \caption{Rewrite steps to prove \coq{hylo_fusion_l} in our mechanisation. The
steps are exactly the same that woudld be required in a manual pen-and-paper
proof.}
\label{fig:proof-fusion-l}
\end{figure}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

  % rewrite hylo_univ, fmap_comp, compA, <- E2,
  %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

>>>>

The remainder of the paper is structured as follows:
\begin{itemize}
  \item In Section~\ref{sec:rec-schemes} we give an overview about recursion
schemes.
  \item In Section~\ref{sec:containers} we formalise the type of container
functors ensuring the presence of least and greatest fixed-points for functors
and suitably adapted for program extraction.
  \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
hylomorphisms, as well as proving their uniqueness properties in Coq.
  \item In Section~\ref{sec:examples} we use the framework to formalise examples
of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
Furthermore we verify the short-cut fusion optimisation and show the extracted
optimised code to OCaml.
\end{itemize}

@@

\subsubsection{Haskell Implementation of Hylomorphisms}
which is very useful in list processing. As a fold, The frequently used function
map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}
 and can be

\begin{minted}{haskell}
hylo :: Functor f => (c -> f c) -> (f a -> a) -> c -> a
hylo c a = a . fmap (hylo c a) . c
\end{minted}

 As an example, we the quicksort program can be written as an hylomorphism

% quicksort :: [Int] -> [Int]
% quicksort (x:xs) = let (xs, x, ys) = partition x xs
%                in (quicksort xs) ++ [x] ++ (quicksort ys)
%   where
%     partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r > x])

\begin{minted}{haskell}
data TreeF a k = Leaf | Branch k a k deriving Functor
\end{minted}

\begin{minted}{haskell}
divide :: [Int] -> TreeF Int [Int]
divide [] = Leaf
divide (p:xs) = Branch [l | l <- xs, l < p] p [r | r <- xs, r > p]
\end{minted}

\begin{minted}{haskell}
conquer :: TreeF Int [Int] -> [Int]
conquer Leaf = []
conquer (Branch ls x rs) = ls ++ [x] ++ rs
\end{minted}
\begin{minted}{haskell}
quicksort' = hylo divide conquer
\end{minted}

\subsubsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the uniqueness property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

\noindent
As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
hylo : forall F `{setoid A} `{setoid B}, Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_uniq (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from this uniqueness property:
\begin{coqcode}
Lemma hylo_fusion_l
  (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
  : f2 \o g1 =e g2 \o fmap f2 ->
    f2 \o hylo g1 h1 =e hylo g2 h1.

@@

apply hylo_uniq.
       (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
rewrite fmap_comp.
       (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
rewrite [...]. (* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
rewrite <- E2.
       (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1      *)
rewrite [...]. (* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
rewrite <- hylo_unroll.
       (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1                             *)
\end{coqcode}
  \caption{Rewrite steps to prove \coq{hylo_fusion_l} in our mechanisation. The
steps are exactly the same that woudld be required in a manual pen-and-paper
proof.}
\label{fig:proof-fusion-l}
\end{figure}
% Qed.
%   rewrite hylo_uniq, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

  % rewrite hylo_uniq, fmap_comp, compA, <- E2,
  %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
